---
title: "Figure_1_problem"
output: html_document
date: "2024-05-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(snpR)
library(ggplot2)
library(tidyverse)
library(khroma)
library(cowplot)
library(ggpubr)
```

This script will produce Figure 1, which demonstrates the high-FST/ascertainment bias issue in PCA and other assignment/clustering approaches.

# Pull in data

We use monarch data from [this paper](doi.org/10.1111/mec.16592) to illustrate the overfitting problem created by taking high $F_{ST}$ loci when doing clustering/assignment testing. Specifically, we will use samples from North America, divided randomly into two populations.

```{r}
set.seed(1232321)

## monarchs clean up
monarchs <- readRDS("../../data/monarch_nomaf.RDS")
colnames(sample.meta(monarchs)) <- c("sampID", "pop", ".sample.id")
sample.meta(monarchs)$sampID <- gsub("\\.", "_", sample.meta(monarchs)$sampID)
sample.meta(monarchs) <- sample.meta(monarchs)[,1:2]

## randomly assign A-D to panmictic north america samples 
monarchs_random_allsnps <- filter_snps(monarchs[pop = "NAM"], maf = 0.05, min_ind = .75, min_loci = .75)
sample.meta(monarchs_random_allsnps)$pop <- sample(c("A", "B", "C", "D"), nsamps(monarchs_random_allsnps), TRUE)

## save for later
saveRDS(monarchs_random_allsnps, "../../data/monarchs_random.RDS")
```

# Grab high $F_{ST}$ loci

Next, we will grab the highest (top 5%) $F_{ST}$ loci the dataset.

```{r}
#### Monarch data for testing ascertainment bias using highly and non differentiated populations
get_high_fst <- function(rds, percent_highest){
  rds <- calc_global_fst(rds, "pop")
  fst <- get.snpR.stats(rds, "pop", "fst")
  high_fst <- which(fst$pairwise$fst >= quantile(fst$pairwise$fst, percent_highest, na.rm = TRUE))
  high_fst <- snpR:::.paste.by.facet(fst$pairwise, c("group", "position"), "_")[high_fst]
  high_fst <- match(high_fst,
                    snpR:::.paste.by.facet(snp.meta(rds), c("group", "position"), "_"))
  high_fst_rds <- rds[high_fst,]
  return(high_fst_rds)
}

monarchs_random_highfst <- get_high_fst(monarchs_random_allsnps, 0.95)
saveRDS(monarchs_random_highfst, "../../data/monarchs_random_highfst.RDS")
```

# Generate PCAs for figure
```{r}
## Set colors
colours <- color("batlow")
monarch_colors <- colours(4, range=c(0.1,0.8))


# plot PCAs for figure 1 
pca_all  <- plot_clusters(monarchs_random_allsnps, "pop", "pca", alt.palette = monarch_colors)

pca_all$plots$pca <- pca_all$plots$pca + 
  guides(color=guide_legend(title = stringr::str_wrap("Randomly Assigned Population", width =20)))

pca_high <- plot_clusters(monarchs_random_highfst, "pop", "pca", alt.palette = monarch_colors)

```

# Generate STUCTURE plots

```{r}
## plot STRUCTURE for fig 1 
setwd("../../results/structure/all_sites_structure_res/")
structure_allsnps <- plot_structure("merged", k = 4, facet = sample.meta(monarchs_random_allsnps)$pop, alt.palette = monarch_colors[c(1,3,2,4)], facet.order = c("A", "B", "C", "D"))

setwd("../high_fst_structure_res/")
structure_highfst <- plot_structure("merged", k = 4, facet = sample.meta(monarchs_random_highfst)$pop, alt.palette = monarch_colors[c(1,3,2,4)], facet.order = c("A", "B", "C", "D"))

setwd("../../../scripts/figure1_scripts")

```

# Generate Rubias plots
```{r}
## plot RUBIAS results for figure 1 
rubias_all  <- read.table("../../results/rubias/monarchs_random_rubias.txt", header = 1)
rubias_high <- read.table("../../results/rubias/monarchs_random_highfst_rubias.txt", header = 1)

p_rubias_all <- ggplot(rubias_all) +
  geom_bar(aes(fill=inferred_collection, x=collection), position="fill", width =0.975) + 
  labs(x= "Randomly Assigned Population",  y="Proportion of Individuals Assigned") +
  scale_fill_manual(values=monarch_colors[3:4]) +
  theme_bw() + 
  guides(fill=guide_legend(title = stringr::str_wrap("Inferred Population", width =20)))
  
p_rubias_high <- ggplot(rubias_high) +
  geom_bar(aes(fill=inferred_collection, x=collection), position="fill", width =0.975) + 
  labs(x= "Randomly Assigned Population",  y="Proportion of Individuals Assigned") +
  scale_fill_manual(values=monarch_colors)+ theme_bw() + 
  guides(fill=guide_legend(title = stringr::str_wrap("Inferred Population", width =20)))
```

# Combine and save
```{r}

top_lg <- get_legend(pca_all$plots$pca + 
                       theme(legend.title = element_text(size=14),
                             legend.text = element_text(size=12)))
mid_lg <- get_legend(structure_allsnps$plot + 
                       theme(legend.title = element_text(size=14),
                             legend.text = element_text(size=12)))
btm_lg <- get_legend(p_rubias_high + 
                       theme(legend.title = element_text(size=14),
                             legend.text = element_text(size=12)))

figure_1 <- plot_grid(
  pca_all$plots$pca + 
    guides(color="none") + 
    ggtitle("All SNPs") + 
    theme(plot.title = element_text(size=20), 
          axis.text = element_text(angle=0, size=12), 
          axis.title = element_text(size=14)), 
  
  pca_high$plots$pca + guides(color="none") + 
    ggtitle(expression('Top' * ' ' * '5%' * ' ' *italic("F")[ST])) + 
    theme(plot.title = element_text(size=20), 
          axis.text = element_text(angle=0, size=12), 
          axis.title = element_text(size=14)), 
  top_lg,
  
  structure_allsnps$plot + 
    guides(fill="none", color="none") + 
    xlab("Randomly Assigned Population") + 
    theme(axis.text = element_text(angle=0, size=12), 
          axis.text.x = element_text(angle=0),
          axis.title = element_text(size=14)),
  
  structure_highfst$plot + 
    guides(fill="none", color="none") + 
    xlab("Randomly Assigned Population") + 
    theme(axis.text = element_text(size=12), 
          axis.text.x = element_text(angle=0),
          axis.title = element_text(size=14)),
  mid_lg,
  
  p_rubias_all + guides(fill="none") + 
    scale_x_discrete(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0)) + 
    theme(axis.text = element_text(angle=0, size=12), 
          axis.title = element_text(size=14)),
  
  p_rubias_high + guides(fill="none") + 
    scale_x_discrete(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0)) +
    theme(axis.text = element_text(angle=0, size=12),
          axis.title = element_text(size=14)), 
  
  btm_lg,
  ncol = 3, axis = "tr", align="hv", rel_widths =c(1,1,.3)
)
figure_1
save_plot("../../results/Figure1.pdf", figure_1,  base_height = 11, base_width = 15)

```

# Other plots for SI

## Generate DAPC, UMAP, tSNE plots for the random data

Next, generate plots using the various clustering approaches. All of these will go to the supplementary material.

For the NAM DAPC, use the following: 6 PCs, 4 clusters, 3 discriminants

```{r}
generate_clusters <- function(rds,
                              npca,
                              nclust,
                              ndisc){
  plots <- plot_clusters(rds, "pop", plot_type = c("dapc", "tsne", "umap"),
                         dapc_clustering_npca = npca, dapc_clustering_nclust = nclust, 
                         dapc_npca = npca, dapc_ndisc = ndisc,
                         dapc_clustering_max_n_clust = NULL,
                         simplify_output = TRUE)
  return(plots)
}


nam_rand_plots <- generate_clusters(monarchs_random_highfst, 6, 4, 2)

```

## sNMF and snapclust
```{r}
snmf <- plot_structure(monarchs_random_highfst, "pop", 
                        method = "snmf", 
                        structure_path = "/usr/bin/structure.exe",
                        k = 2:4, iterations = 500, burnin = 100)

snapclust <- plot_structure(monarchs_random_highfst, "pop", 
                        method = "snapclust", 
                        structure_path = "/usr/bin/structure.exe",
                        k = 2:4, iterations = 500, burnin = 100)
```

